Sorting algorithms: how are they affected by memory faults?

### 2019-12-23 paper/DataAnalysis.tex
Revisao 01 do paper

In this section, we present our results after the execution of sorting algorithms over the input files. We analyzed only the dependent variables \textit{percentage of the largest subarray size ((\%LSS)} and \textit{percentage of unordered elements quantity (\%UEQ)}. These variables, because they are a percentage value, already were normalized (i.e., the same order of magnitude) related to dependent variable \textit{array size}.

@@

After the Shapiro-Wilk test, we determine that only the dependent variable \%UEQ (\textit{percentage of unordered elements quantity}) has a normal distribution related to mean for all algorithms. This situation occurs considering all possible combinations between independent variables: \textit{sorting algorithm}, \textit{probability of failure} and \textit{array size}.

Based on these results, we choose to test just the hypothesis associated with the dependent variable \%UEQ. We make use of the ANOVA method to test the hypothesis, and this method is premised the normal distribution of variables values.

@@

\begin{table}[H]
    \parbox{.45\linewidth}{
        \caption{Shapiro-Wilk test for \%LSS with \textit{probability of failure} of 5\% and \textit{array size} of 1000.}
        \begin{center}
        \begin{tabular}{|c|c|c|}
        \hline
        \textbf{Algorithm} & \textbf{W} & \textbf{p-value} \\
        \hline
        Bubblesort & 0.891642 & 0.00052 \\
        \hline
        Insertion Sort & 0.921918 & 0.3000 \\
        \hline
        Mergesort & 0.878308 & 0.0025 \\
        \hline
        Quicksort & 0.917583 & 0.0232 \\
        \hline
        \end{tabular}
        \label{table-shapiro-test-lss-005-1000}
        \end{center}
    }
    \hfill
    \parbox{.45\linewidth}{
        \caption{Shapiro-Wilk test for \%UEQ with \textit{probability of failure} of 5\% and \textit{array size} of 1000.}
        \begin{center}
        \begin{tabular}{|c|c|c|}
        \hline
        \textbf{Algorithm} & \textbf{W} & \textbf{p-value} \\
        \hline
        Bubblesort & 0.936459 & \textbf{0.0730} \\
        \hline
        Insertion Sort & 0.931681 & \textbf{0.0544} \\
        \hline
        Mergesort & 0.961920 & \textbf{0.3465} \\
        \hline
        Quicksort & 0.963954 & \textbf{0.3892} \\
        \hline
        \end{tabular}
        \label{table-shapiro-test-ueq-005-1000}
        \end{center}
    }
\end{table}

@@

\begin{table}[H]
    \caption{Hypothesis 3 ANOVA results}
    \begin{center}
    \begin{tabular}{|c|c|c|c|c|}
    \hline
    \multirow{2}{*}{\hfill}&\multicolumn{4}{|c|}{\textbf{Bubblesort}} \\
    \cline{2-5}
    &\textbf{\textit{sum\_sq}} & \textbf{\textit{df}} & \textbf{\textit{F}} & \textbf{\textit{PR(>F)}} \\
    \hline
    size\_of\_array & 0.019548 & 1.0 & 0.01350 & \textbf{0.90759} \\
    \hline
    probability\_of\_failure & 3590.644437 & 1.0 & 2479.71637 & \textbf{7.570027e-137} \\
    \hline
    residual & 385.169623 & 266.0 & &  \\
    \hline
    \multirow{2}{*}{\hfill}&\multicolumn{4}{|c|}{\textbf{Mergesort}} \\
    \cline{2-5}
    &\textbf{\textit{sum\_sq}} & \textbf{\textit{df}} & \textbf{\textit{F}} & \textbf{\textit{PR(>F)}} \\
    \hline
    size\_of\_array & 1784.873256 & 1.0 & 383.585800 & \textbf{1.702008e-53} \\
    \hline
    probability\_of\_failure & 13009.963175 & 1.0 & 2795.961630 & \textbf{3.801570e-143} \\
    \hline
    residual & 1237.731651 & 266.0 & &  \\
    \hline
    \multirow{2}{*}{\hfill}&\multicolumn{4}{|c|}{\textbf{Quicksort}} \\
    \cline{2-5}
    &\textbf{\textit{sum\_sq}} & \textbf{\textit{df}} & \textbf{\textit{F}} & \textbf{\textit{PR(>F)}} \\
    \hline
    size\_of\_array & 121.551362 & 1.0 & 32.964196 & \textbf{2.556033e-08} \\
    \hline
    probability\_of\_failure & 5626.955128 & 1.0 & 1526.005558 & \textbf{3.457996e-112} \\
    \hline
    residual & 980.841817 & 266.0 & &  \\
    \hline
    \multirow{2}{*}{\hfill}&\multicolumn{4}{|c|}{\textbf{Insertion sort}} \\
    \cline{2-5}
    &\textbf{\textit{sum\_sq}} & \textbf{\textit{df}} & \textbf{\textit{F}} & \textbf{\textit{PR(>F)}} \\
    \hline
    size\_of\_array & 309.843991 & 1.0 & 75.905868 & \textbf{3.236586e-16} \\
    \hline
    probability\_of\_failure & 6806.075083 & 1.0 & 1667.358570 & \textbf{1.414758e-116} \\
    \hline
    residual & 1085.798822 & 116.0 & & \\
    \hline
    \end{tabular}
    \label{tab-hypothesis3-testing}
    \end{center}
\end{table}

@@

For Bubblesort algorithm, however, there is no confidence to reject the null hypothesis ($H_{0}$). Analyzing the result, we can affirm with 95\% of confidence that the value of \textit{probability of failure} variable had a significant impact on the value of the dependent variable \%UEQ. But we can't affirm the same related to independent variable \textit{array size}. These results demonstrate that, for Bubblesort algorithm, the array sizes used in this experiment were not decisive to define values concerning the percentage of unordered elements quantity.

@@

In this section, we present an analysis of overall performance related to dependent variable \textit{percentage of unordered elements quantity}. To help this analysis, we show below 3 bar graphs, each of these for a different value of \textit{probability of failure} independent variable. These graphs shows informations about values of \%UEQ variable grouped by \textit{array size} and \textit{sorting algorithm} variables. Figures \ref{fig-barplot-pof-001}, \ref{fig-barplot-pof-002}, and \ref{fig-barplot-pof-005} below exhibit these graphs.

The bars in each graph represents the sample mean value for variable \%UEQ inside each group. Above each bar, there is a vertical line that indicates the lower and upper limits of a confidence interval for a significance level of 5\% ($\alpha = 0.05$).

@@

Considering the data in the graphs, we can conclude that Bubblesort generates lower mean value for variable \%UEQ for all defined combinations (\textit{probability of failure} X \textit{array size} X \textit{sorting algorithm}). These data demonstrate that Bubblesort was the less affected algorithm by the memory faults simulated in this experiment. We believe that this fact happens because of the operation of the algorithm itself, where all elements order are verified each iteration, causing an element incorrectly positioned in an iteration (because of a memory fault) can have its location corrected in an upcoming iteration.

Related to Bubblesort yet, we could identify that despite different sizes of input arrays (100, 1000, and 10000), there wasn't a significant impact on the mean value obtained for \%UEQ variable, proving what we verify with ANOVA when doing hypothesis 3 testing for this algorithm.

>>>>

In this section, we present our results after the execution of sorting algorithms over the input files. We analyzed only two of the four dependent variables, which were \textit{percentage of the largest subarray size (\%LSS)} and \textit{percentage of unordered elements quantity (\%UEQ)}. These variables, because they are a percentage value, already were normalized (i.e., the same order of magnitude) related to dependent variable \textit{array size}.

@@

After the Shapiro-Wilk test, we determine that only the dependent variable \%UEQ (\textit{percentage of unordered elements quantity}) has a normal distribution related to mean for all algorithms. This situation occurred considering all possible combinations between independent variables: \textit{sorting algorithm}, \textit{probability of failure} and \textit{array size}.

Based on these results, we choose to test just the hypothesis associated with the dependent variable \%UEQ. We made use of the ANOVA method to test the hypothesis, and this method is premised the normal distribution of variables values.

@@

\begin{table}[H]
    \parbox{.45\linewidth}{
        \caption{Shapiro-Wilk test for \%LSS with \textit{probability of failure} of 5\% and \textit{array size} of 1000.}
        \begin{center}
        \begin{tabular}{|c|c|c|}
        \hline
        \textbf{Algorithm} & \textbf{W} & \textbf{p-value} \\
        \hline
        Bubblesort & 0.891642 & 0.0052 \\
        \hline
        Insertion Sort & 0.921918 & 0.0300 \\
        \hline
        Mergesort & 0.878308 & 0.0025 \\
        \hline
        Quicksort & 0.917583 & 0.0232 \\
        \hline
        \end{tabular}
        \label{table-shapiro-test-lss-005-1000}
        \end{center}
    }
    \hfill
    \parbox{.45\linewidth}{
        \caption{Shapiro-Wilk test for \%UEQ with \textit{probability of failure} of 5\% and \textit{array size} of 1000.}
        \begin{center}
        \begin{tabular}{|c|c|c|}
        \hline
        \textbf{Algorithm} & \textbf{W} & \textbf{p-value} \\
        \hline
        Bubblesort & 0.936459 & \textbf{0.0730} \\
        \hline
        Insertion Sort & 0.931681 & \textbf{0.0544} \\
        \hline
        Mergesort & 0.961920 & \textbf{0.3465} \\
        \hline
        Quicksort & 0.963954 & \textbf{0.3892} \\
        \hline
        \end{tabular}
        \label{table-shapiro-test-ueq-005-1000}
        \end{center}
    }
\end{table}

@@

\begin{table}[H]
    \caption{Hypothesis 3 ANOVA results}
    \begin{center}
    \begin{tabular}{|c|c|c|c|c|}
    \hline
    \multirow{2}{*}{\hfill}&\multicolumn{4}{|c|}{\textbf{Bubblesort}} \\
    \cline{2-5}
    &\textbf{\textit{sum\_sq}} & \textbf{\textit{df}} & \textbf{\textit{F}} & \textbf{\textit{PR(>F)}} \\
    \hline
    size\_of\_array & 0.019548 & 1.0 & 0.01350 & \textbf{0.90759} \\
    \hline
    probability\_of\_failure & 3590.644437 & 1.0 & 2479.71637 & \textbf{7.570027e-137} \\
    \hline
    residual & 385.169623 & 266.0 & &  \\
    \hline
    \multirow{2}{*}{\hfill}&\multicolumn{4}{|c|}{\textbf{Mergesort}} \\
    \cline{2-5}
    &\textbf{\textit{sum\_sq}} & \textbf{\textit{df}} & \textbf{\textit{F}} & \textbf{\textit{PR(>F)}} \\
    \hline
    size\_of\_array & 1784.873256 & 1.0 & 383.585800 & \textbf{1.702008e-53} \\
    \hline
    probability\_of\_failure & 13009.963175 & 1.0 & 2795.961630 & \textbf{3.801570e-143} \\
    \hline
    residual & 1237.731651 & 266.0 & &  \\
    \hline
    \multirow{2}{*}{\hfill}&\multicolumn{4}{|c|}{\textbf{Quicksort}} \\
    \cline{2-5}
    &\textbf{\textit{sum\_sq}} & \textbf{\textit{df}} & \textbf{\textit{F}} & \textbf{\textit{PR(>F)}} \\
    \hline
    size\_of\_array & 121.551362 & 1.0 & 32.964196 & \textbf{2.556033e-08} \\
    \hline
    probability\_of\_failure & 5626.955128 & 1.0 & 1526.005558 & \textbf{3.457996e-112} \\
    \hline
    residual & 980.841817 & 266.0 & &  \\
    \hline
    \multirow{2}{*}{\hfill}&\multicolumn{4}{|c|}{\textbf{Insertionsort}} \\
    \cline{2-5}
    &\textbf{\textit{sum\_sq}} & \textbf{\textit{df}} & \textbf{\textit{F}} & \textbf{\textit{PR(>F)}} \\
    \hline
    size\_of\_array & 309.843991 & 1.0 & 75.905868 & \textbf{3.236586e-16} \\
    \hline
    probability\_of\_failure & 6806.075083 & 1.0 & 1667.358570 & \textbf{1.414758e-116} \\
    \hline
    residual & 1085.798822 & 116.0 & & \\
    \hline
    \end{tabular}
    \label{tab-hypothesis3-testing}
    \end{center}
\end{table}

@@

For Bubblesort algorithm, however, there is no confidence to reject the null hypothesis ($H_{0}$). Analyzing the result, we can affirm with 95\% of confidence that the value of \textit{probability of failure} variable had a significant impact on the value of the dependent variable \%UEQ. But we can not affirm the same related to independent variable \textit{array size}. These results demonstrate that, for the Bubblesort algorithm, the array sizes used in this experiment were not decisive to define values concerning the percentage of unordered elements quantity.

@@

In this section, we present an analysis of overall performance related to dependent variable \textit{percentage of unordered elements quantity}. To help this analysis, we show below 3 bar graphs, each of these for a different value of \textit{probability of failure} independent variable. These graphs show information about values of \%UEQ variable grouped by \textit{array size} and \textit{sorting algorithm} variables. Figures \ref{fig-barplot-pof-001}, \ref{fig-barplot-pof-002}, and \ref{fig-barplot-pof-005} below exhibit these graphs.

The bars in each graph represent the sample mean value for variable \%UEQ inside each group. Above each bar, there is a vertical line that indicates the lower and upper limits of a confidence interval for a significance level of 5\% ($\alpha = 0.05$).

@@

Considering the data in the graphs, we can conclude that Bubblesort generates lower mean value for variable \%UEQ for all defined combinations (\textit{probability of failure} X \textit{array size} X \textit{sorting algorithm}). These data demonstrate that Bubblesort was the less affected algorithm by the memory faults simulated in this experiment. We believe that this fact happened because of the operation of the algorithm itself, where all elements order are verified each iteration, causing an element incorrectly positioned in an iteration (because of a memory fault) can have its location corrected in an upcoming iteration.

Related to Bubblesort yet, we could identify that despite different sizes of input arrays (100, 1000, and 10000), there was not a significant impact on the mean value obtained for \%UEQ variable, proving what we verify with ANOVA when doing hypothesis 3 testing for this algorithm.

### 2019-12-23 paper/ExperimentalSetup.tex
Revisao 01 do paper

\begin{table}[H]
    \caption{Values of the independent variables.}
    \begin{center}
    \begin{tabular}{|c|c|}
    \hline
    \textbf{Variable} & \textbf{Values} \\
    \hline
    Probability of failure & 1\%, 2\% and 5\% \\
    \hline
    Array size & 100, 1000 and 10000 \\
    \hline
    Sorting algorithm & Bubblesort, Quicksort, Mergesort and Insertion sort \\
    \hline
    \end{tabular}
    \label{table-independent-variables-values}
    \end{center}
\end{table}

@@

\begin{itemize}
    \item \textit{Probability of Failure}: the first number of the sequence (\texttt{0.01}) is the probability of memory failure when sorting;
    \item \textit{Sequence size}: the second number (\texttt{100}) means the size of the integers sequence used by sorting;
    \item \textit{Sequence}: the rest of the numbers indicates the sequence itself.
\end{itemize}

With this input data, we ran, for each one of these, all four algorithms considered in this study. The sorting algorithms used already existed. For example, using all 270 input files, we ran bubblesort, creating 270 output files, and so on for the other algorithms. At the end of executions, we get a total of 1080 output files. An output file look like shown in Figure \ref{fig-output-file-example}:

@@

The output file gives four essential data, as enumerated below:
\begin{itemize}
    \item \textit{[1]}: the original sequence of integers contained in the input file;
    \item \textit{[2]}: the sequence processed by the sorting algorithm under the memory fault model;
    \item \textit{[3]}: the sequence sorted correctlty;
    \item \textit{[4]}: the size of the largest sorted subsequence in [2]. This number can be interpreted as the quality of sorting. As higher, most successful was the sorting operation.
\end{itemize}

After generating the dataset, we developed a Python script that reads the 1080 output files and produces a single CSV file (first lines showed in Figure \ref{fig-output-csv-file-example} below). This file was used to run the statistical tests, and contains the following columns:
\begin{itemize}
    \item \textit{algorithm}: the algorithm used to sort the array;
    \item \textit{probability\_of\_failure}: the probability of failure used when sorting;
    \item \textit{size\_of\_array}: the size of the array to be sorted;
    \item \textit{largest\_sorted\_subarray}: the largest sorted subarray after sorting;
    \item \textit{k\_unordered\_sequence}: number of unordered sequence after sorting.
    \item \textit{percentage\_k\_unordered}: percentage of unordered sequence after sorting related to original array;
    \item \textit{percentage\_largest\_sorted\_subarray}: percentage of largest sorted subarray after sorting related to original array.
\end{itemize}

\begin{verbbox}[\mbox{}]
algorithm;probability_of_failure;size_of_array;largest_sorted_subarray;
k_unordered_sequence;percentage_k_unordered;
percentage_largest_sorted_subarray
quick;0.01;100;35;4;4.00;35.00
quick;0.01;100;36;8;8.00;36.00
quick;0.01;100;31;5;5.00;31.00
quick;0.01;100;20;6;6.00;20.00
quick;0.01;100;31;5;5.00;31.00
\end{verbbox}

>>>>

\begin{table}[H]
    \caption{Values of the independent variables.}
    \begin{center}
    \begin{tabular}{|c|c|}
    \hline
    \textbf{Variable} & \textbf{Values} \\
    \hline
    Probability of failure & 1\%, 2\% and 5\% \\
    \hline
    Array size & 100, 1000 and 10000 \\
    \hline
    Sorting algorithm & Bubblesort, Quicksort, Mergesort and Insertionsort \\
    \hline
    \end{tabular}
    \label{table-independent-variables-values}
    \end{center}
\end{table}

@@

\begin{itemize}
    \item \textit{Probability of Failure}: the first number of the sequence (\texttt{0.01}) is the probability of memory failure when sorting;
    \item \textit{Sequence size}: the second number (\texttt{100}) means the size of the integers sequence used by sorting;
    \item \textit{Sequence}: the rest of the numbers represents a list of \textit{n} positive integers that will be sorted.
\end{itemize}

With this input data, we ran, for each one of these, all four algorithms considered in this study. The sorting algorithms used already existed. For example, using all 270 input files, we ran bubblesort, creating 270 output files, and so on for the other algorithms. At the end of executions, we get a total of 1080 output files. An output file looks like shown in Figure \ref{fig-output-file-example}:

@@

The output file gives four essential data, as enumerated below:
\begin{itemize}
    \item \textit{1)} the original sequence of integers contained in the input file;
    \item \textit{2)} the sequence processed by the sorting algorithm under the memory fault model;
    \item \textit{3)} the sequence sorted correctly;
    \item \textit{4)} the size of the largest sorted subsequence in item \textit{2}. This number can be interpreted as a quality measure. As higher, most successful was the sorting operation.
\end{itemize}

After generating the dataset, we developed a Python program that reads the 1080 output files and produces a single CSV file (first lines showed in Figure \ref{fig-output-csv-file-example}). This file contains the columns listed below and was used to perform the exploratory data analysis and to run the statistical tests.
\begin{itemize}
    \item \textit{sorting\_algorithm}: the algorithm used to sort the array;
    \item \textit{probability\_of\_failure}: the probability of failure used when sorting;
    \item \textit{array\_size}: the size of the sorted array;
    \item \textit{largest\_subarray\_size}: the largest sorted subarray after sorting;
    \item \textit{unordered\_elements\_quantity}: quantity of unordered sequence elements after sorting.
    \item \textit{perc\_unordered\_elements\_quantity}: percentage of unordered elements after sorting related to original array;
    \item \textit{perc\_largest\_subarray\_size}: percentage of largest sorted subarray after sorting related to original array.
\end{itemize}

\begin{verbbox}[\mbox{}]
sorting_algorithm;probability_of_failure;array_size;largest_subarray_size;
unordered_elements_quantity;perc_unordered_elements_quantity;
perc_largest_subarray_size
quick;0.01;100;35;4;4.00;35.00
quick;0.01;100;36;8;8.00;36.00
quick;0.01;100;31;5;5.00;31.00
quick;0.01;100;20;6;6.00;20.00
quick;0.01;100;31;5;5.00;31.00
\end{verbbox}
